```{r}
library(pacman)
p_load(data.table, dplyr, DESeq2, tximport, DT, foreach, clusterProfiler, org.Hs.eg.db, rtracklayer, 
ComplexHeatmap, RColorBrewer, ggplot2, DESeq2, ggrepel) 
options(scipen=999)
knitr::opts_chunk$set(fig.width=6, fig.height=6)
dir.create(outdatadir, recursive=TRUE)
pthres <- 0.05
lfc <- log2(2)
source("code/common.R")
dir.create(paste0("report/out_html/code/", basename(dirname(outdatadir)), "/", basename(outdatadir), "/pval/"), recursive=TRUE)
dir.create(paste0("report/out_html/code/", basename(dirname(outdatadir)), "/", basename(outdatadir), "/pvalOnly/"), recursive=TRUE)
```

Samples used in this page: 

```{r}
tmp <- readRDS(paste0("output/code/lncRNA/DataQC/", "data2DEG.RDS"))
txi <- tmp[[1]]
sampleInfo <- tmp[[2]]

controle <- "control"#strsplit(basename(outdatadir), "_VS_") %>% sapply(., `[`, 2)
samplas <- "Batimastat" #strsplit(basename(outdatadir), "_VS_") %>% sapply(., `[`, 1)
samples2do <- c("T1H1", "T1H2", "T1cH1", "T1cH2" )
sampleInfo <- sampleInfo[Barcode %in% samples2do, ] 
sampleInfo$Group <- sampleInfo$Treatment

DT::datatable(sampleInfo[, .(sample, replicate)], class = 'cell-border stripe')

tmp <- foreach(i=1:3) %do% {
  txi[[i]] <- txi[[i]][, colnames(txi[[i]]) %in% sampleInfo$sample]  
}
sampleInfo <- sampleInfo[order(match(sample, colnames(txi[[1]])))]
sampleInfo$group <- sampleInfo$Group
stopifnot(all(colnames(txi[[1]]) == sampleInfo$Barcode))
```

# Results {.tabset}

```{r}
get_DDS <- function(txi, sampleTable, fonas, pvalue_th=0.05, LFC_th=log(2)) {
  #x <- countai 
  #sampleTable <- sampleInfo   
  #background <- "control"
  dds_o <-  DESeqDataSetFromTximport(
      txi,
      sampleTable,      
      design = ~Group
      )

  dds_o$Group <- relevel(dds_o$Group, ref = fonas)  
  dds <- DESeq(dds_o, quiet=TRUE, fitType="parametric")
  coef_name <- resultsNames(dds)[grepl("Group_", resultsNames(dds))]
  res <- results(dds, 
                 alpha=pvalue_th, 
                 lfcThreshold=LFC_th)
  vsd <- vst(dds, blind=FALSE)
  vsd_blind <- vst(dds, blind=TRUE)
  dds_shrunken <- lfcShrink(dds, 
                            coef = coef_name, 
                            type = "apeglm")  
  return(list(dds=dds, 
              dds_o=dds_o, 
              res=res, 
              vsd=vsd, 
              vsd_blind=vsd_blind, 
              dds_shrunken=dds_shrunken))
}

dds_list <- get_DDS(txi, sampleInfo, controle, pthres, lfc)
dge_id <- dds_list[["res"]] %>% as.data.table(., keep.rownames=TRUE) %>% .[padj <= pthres & abs(log2FoldChange) >= lfc, rn] 
saveRDS(dds_list, paste0(outdatadir, controle, "_vs_", samplas, ".RDS"))
countai <- counts(dds_list[["dds"]]) %>% as.data.table
vst_counts <- assay(dds_list[["vsd"]]) %>% as.data.table(., keep.rownames=TRUE) 
```

## Quality plots 

This part include PCA and correlations graph. 
In the PCA we hope to see segregation between our sample groups on the first PC. We also hope to see that first PC explain was majority of data variability. 

Most probably we will see a trend, but results will not be ideal. Most probably we will see some bias in biological replicates. 

```{r}
get_PCA(countai, sampleInfo)
```

Correlation plot (pearson correlation) should show higher correlation within a group (in comparison to correlation between groups). 


```{r}
ht <- get_cor(countai, sampleInfo)
draw(ht, annotation_legend_side = "right", heatmap_legend_side = "bottom", legend_grouping = "original")
```

Second PCA plot was made using normalized counts (VST transformation). This one should look better. 

```{r}
get_PCA(assay(dds_list[["vsd"]]), sampleInfo, "PCA on VST transformed counts")
```

## DEG plots

In this part we have a heatmap, a Volcano plot, an interactive MA plot and results table (with all genes being analysed). 

A Volcano plot shows relationship between adjusted p-value and log2fold change. 

```{r}
get_volcano(dds_list[["res"]], pthres, lfc)
```

A heatmap is a classical way to visualize differentially expressed genes.

```{r}
get_heatmap(dge_id, vst_counts, paste0(controle, " vs. ", samplas))
```


```{r}
lncRNA_types <- readRDS("output/code/General/lncRNA_list/lncRNA_types.RDS")

lncRNA_types %>% 
  as.data.table() %>% 
  .[gene_id %in% dge_id, type] %>% 
  table() %>% 
  as.data.table() %>% 
  setnames(., c("Type", "amount")) %>% 
  ggplot(aes(x=Type, y=amount, fill=Type, label=amount)) +
    geom_col() +
    theme_bw() +
    geom_label() +
    #geom_label_repel() +
    scale_fill_brewer(palette="Dark2") +
    coord_flip() +
    theme(legend.position="bottom")
```

## DEG plots II  

Labai panaši analizė, bet pridėtas fold change stabilizavimas (LFC shrinkage) naudojant apeglm. 


```{r}
get_volcano(dds_list[["dds_shrunken"]], pthres, lfc)
dge_id_shrunken <- dds_list[["dds_shrunken"]] %>% 
  as.data.table(., keep.rownames=TRUE) %>% 
  .[abs(log2FoldChange) >= lfc & padj <= pthres, rn]
```

A heatmap is a classical way to visualize differentially expressed genes.

```{r}
get_heatmap(dge_id_shrunken, vst_counts, paste0(controle, " vs. ", samplas))
```

```{r, fig.height=4}
lncRNA_types %>% 
  as.data.table() %>% 
  .[gene_id %in% dge_id_shrunken, type] %>% 
  table() %>% 
  as.data.table() %>% 
  setnames(., c("Type", "amount")) %>% 
  ggplot(aes(x=Type, y=amount, fill=Type, label=amount)) +
    geom_col() +
    theme_bw() +
    geom_label() +
    #geom_label_repel() +
    scale_fill_brewer(palette="Dark2") +
    coord_flip() +
    theme(legend.position="bottom")
```
